---
title: "Monthly Expense Report"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
params:
  username: "NA"
  month: "YYYY-MM"
  salary: 0
  goal: 0
  total_expense: 0
  remaining: 0
  transactions: NULL
   
  category_plot: NULL
  date_plot: NULL
  payment_mode_plot: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(dplyr)
library(ggplot2)
library(lubridate)
```

# User Information

**Username:** `r params$username`  
**Report Month:** `r params$month`

# Salary & Savings Summary

| Description        | Amount (Rs.)     |
|--------------------|------------------|
| Monthly Salary     | `r format(params$salary, big.mark = ",")` |
| Savings Goal       | `r format(params$goal, big.mark = ",")` |
| Total Expenses     | `r format(params$total_expense, big.mark = ",")` |
| Remaining Amount   | `r format(params$remaining, big.mark = ",")` |

\newpage

# Transaction Details

```{r transaction_table}
if (nrow(params$transactions) == 0) {
  cat("No transactions found for this month.")
} else {
  knitr::kable(params$transactions, format = "pipe", align = "c")
}
```

\newpage

# Visualizations


## Spending by Category

```{r category_plot, fig.align='center', fig.height=5}
if (!is.null(params$transactions) && nrow(params$transactions) > 0) {
  category_summary <- params$transactions %>%
    group_by(category) %>%
    summarise(total = sum(amount)) %>%
    mutate(percent = round(total / sum(total) * 100, 1),
           label = paste0(category, " (", percent, "%)"))

  ggplot(category_summary, aes(x = "", y = total, fill = label)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y") +
    
    theme_void() +
    theme(legend.title = element_blank())
} else {
  cat("No category data available.")
}
```

\newpage

## Spending by Date

```{r date_plot, fig.align='center', fig.height=5}
if (!is.null(params$transactions) && nrow(params$transactions) > 0) {
  date_summary <- params$transactions %>%
    group_by(date = as.Date(date)) %>%
    summarise(total = sum(amount))

  ggplot(date_summary, aes(x = date, y = total)) +
    geom_line(color = "darkgreen", size = 1.2) +
    geom_point(color = "black") +
    labs( x = "Date", y = "Total (Rs.)") +
    theme_minimal()
} else {
  cat("No date data available.")
}
```

\newpage

## Spending by Payment Mode

```{r payment_mode_plot, fig.align='center', fig.height=5}
if (!is.null(params$transactions) && nrow(params$transactions) > 0) {
  mode_summary <- params$transactions %>%
    group_by(payment_mode) %>%
    summarise(total = sum(amount))

  ggplot(mode_summary, aes(x = reorder(payment_mode, -total), y = total, fill = payment_mode)) +
    geom_col(width = 0.6) +
    labs( x = "Payment Mode", y = "Total (Rs.)") +
    theme_minimal() +
    theme(legend.position = "none")
} else {
  cat("No payment mode data available.")
}

```

\newpage

# Notes

- This report includes all expense records and summaries for the selected month only.  
- Visualizations are generated automatically from the transaction data.

> Report generated on `r Sys.Date()` using *Personal Finance Tracker App*.
